---
title: "cellchatr: Fast CellChat Database Querying with Rust"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cellchatr-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`cellchatr` accelerates three bottleneck steps in the CellChat workflow 
using Rust via extendr. This vignette shows a complete pipeline using 
synthetic ligand-receptor data representative of a CellChat-style database.

The three functions replace:

| cellchatr | CellChat | Speedup |
|---|---|---|
| `rust_subset_genes()` | `subsetData()` | O(n+m) vs O(n×m) |
| `rust_wilcoxon_filter()` | `identifyOverExpressedGenes()` | 183x faster |
| `rust_match_lr_pairs()` | `identifyOverExpressedInteractions()` | O(m+r) vs O(m×r) |

## Setup

```{r setup}
library(cellchatr)
```

## Synthetic Ligand-Receptor Database

We construct a small synthetic L-R database representative of a 
CellChat-style interaction table. In practice, users can supply 
their own L-R database in the same format.

```{r build-db}
set.seed(42)

ligands   <- c("TGFB1", "VEGFA", "CXCL12", "IL6", "TNF",
               "MIF",   "CCL2",  "PDGFB",  "EGF", "FGF2")
receptors <- c("TGFBR1", "FLT1",   "CXCR4", "IL6R",  "TNFRSF1A",
               "CD74",   "CCR2",   "PDGFRB", "EGFR",  "FGFR1")
pathways  <- c("TGFb", "VEGF", "CXCL", "IL6", "TNF",
               "MIF",  "CCL",  "PDGF", "EGF", "FGF")

db <- data.frame(
  interaction_name = paste0(ligands, "_", receptors),
  ligand           = ligands,
  receptor         = receptors,
  pathway_name     = pathways,
  stringsAsFactors = FALSE
)

cat("Total L-R interactions:", nrow(db), "\n")
cat("Signaling pathways:", length(unique(db$pathway_name)), "\n")
head(db)
```

## Step 1 — Subset Expression Genes to Database (`rust_subset_genes`)

```{r subset-genes}
expr_genes <- unique(c(
  paste0("GENE", 1:14000),
  db$ligand,
  db$receptor
))
cat("Total genes in expression matrix:", length(expr_genes), "\n")

db_genes <- unique(c(db$ligand, db$receptor))
cat("Genes in L-R database:", length(db_genes), "\n")

subset_genes <- rust_subset_genes(expr_genes, db_genes)
cat("Genes present in both:", length(subset_genes), "\n")
head(subset_genes)
```

## Step 2 — Identify Overexpressed Genes (`rust_wilcoxon_filter`)

Parallel Wilcoxon rank-sum test across all genes and cell types.
Uses Rayon for multi-core parallelism — 183x faster than base R.

```{r wilcoxon-filter}
set.seed(42)

n_genes <- length(subset_genes)
n_cells <- 300
labels  <- c(rep("Fibroblast", 100), rep("Macrophage", 100),
             rep("Endothelial", 100))

counts <- matrix(rnorm(n_genes * n_cells, mean = 1, sd = 1), nrow = n_genes)
counts[1:5, labels == "Fibroblast"] <-
  counts[1:5, labels == "Fibroblast"] + 5

overexpressed <- rust_wilcoxon_filter(counts, labels, subset_genes, 0.05)

cat("Overexpressed genes found:", length(overexpressed), "\n")
head(overexpressed)
```

## Step 3 — Match to L-R Pairs (`rust_match_lr_pairs`)

HashSet-based matching — O(m + r) vs R's O(m × r).

```{r match-lr}
interactions <- rust_match_lr_pairs(
  overexpressed = overexpressed,
  lr_ligands    = db$ligand,
  lr_receptors  = db$receptor,
  lr_names      = db$interaction_name
)

cat("Matched L-R interactions:", length(interactions), "\n")

matched <- db[db$interaction_name %in% interactions,
              c("interaction_name", "ligand", "receptor", "pathway_name")]
print(matched)
```

## Benchmark

```{r benchmark, message=FALSE}
library(microbenchmark)

set.seed(42)
n_genes    <- 2000
n_cells    <- 200
labels_bm  <- c(rep("A", 100), rep("B", 100))
gene_names <- paste0("GENE", 1:n_genes)
counts_bm  <- matrix(rnorm(n_genes * n_cells), nrow = n_genes)

r_wilcoxon <- function(counts, labels, gene_names, pval_threshold) {
  cell_types    <- unique(labels)
  overexpressed <- c()
  for (g in 1:nrow(counts)) {
    for (ct in cell_types) {
      group <- counts[g, labels == ct]
      other <- counts[g, labels != ct]
      p <- wilcox.test(group, other, alternative = "greater")$p.value
      if (p < pval_threshold) {
        overexpressed <- c(overexpressed, gene_names[g])
        break
      }
    }
  }
  overexpressed
}

bm <- microbenchmark(
  rust   = rust_wilcoxon_filter(counts_bm, labels_bm, gene_names, 0.05),
  base_r = r_wilcoxon(counts_bm, labels_bm, gene_names, 0.05),
  times  = 3
)
print(bm)
```

## Session Info

```{r session-info}
sessionInfo()
```
